Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2014-06-17 omrp_lb_rev.tex
Edits on Comp Efficiency Section
Corrected some typos,
Added more explanations

The increased computational workload from solving overlapping batches raises the question of the computational efficiency of this method to estimate $VG$.
In this setting, computational efficiency may be defined as $[T(\gammab)\var{VG}]$ where $T(\gammab)$ is the computational time.
Because variance reduction is known from (\ref{eq:var_reduct_formula}) we need only examine the computational time, which can be decomposed as $T(\gammab) = t(\gammab)\left(\nb\right)$ where $t(\gammab)$ is the average computation time per batch.
Warm-starting an algorithm may be used to decrease the batch solution time for $\gammab < 1$, leading to a possible increase in computational efficiency.

To investigate this, we assumed that computation time per batch can be approximated as $t(\gammab) \propto \gammab^p$ for some $p \geq 0$.
Results for some values of $p$ are shown in Figure \ref{fig:efficiency_theory}.
Efficiency gains increase with larger exponent $p$, but are positive even for small values of $p$ when $\gammab$ is sufficiently close to 1.
At $p=1$, i.e., for problems where computational time is proportional to the number of samples replaced in the batch, computational efficiency is maximized with maximally overlapping batches.

We tested this theoretical relationship using the Newsvendor problem solved with insertion sort.
The average computation time per batch was found empirically to be well approximated by power law with $p = 0.71$.
The measured computational efficiency is compared to the theoretical result in Figure \ref{fig:efficiency_nv}.

In cases where the exponent $p$ can be estimated the above theoretical method may be used to determine a reasonable value for $\gammab$.
We can see in Figure \ref{fig:efficiency} that for some values of $p$ efficiency gian is nearly flat over a region of $\gammab$.
In this case we recommend choosing a $\gammab$ near the right side of this region.
This choice keeps efficiency high while decreasing computation time and preventing large decreases in efficiency in the event that $p$ was overestimated.

@@

A final question on computational efficiency would be whether OMRP is more efficient than nonoverlapping MRP with a larger number of batches.
In the nonoverlapping case, $\var{VG} \propto \frac{1}{k}$, so decreasing $\var{VG}$ to the level of maximally overlapping batches requires a $50\%$ increase in the number of batches, and thus a $50\%$ increase in computation time.
Again assuming average time per batch can be represented as $T(\gammab) = \gammab^p$, Figure \ref{fig:comp_work_comparison} compares the increase in computational work for OMRP for several values of $p$ with the work increase for MRP to get an equivalent decrease in variance.
We again see that OMRP is advantageous for larger values of $p$.

>>>>

The increased computational workload from solving overlapping batches raises the question of the computational efficiency of this method to estimate $VG$.
In this setting, computational efficiency is defined as $[T(\gammab)\var{VG}]^{-1}$, where $T(\gammab)$ is the computational time spent to generate $VG$ for a given value of $\gammab$.
Because the variance of overlapping estimator relative to the nonoverlapping (MRP) variant is known from (\ref{eq:var_reduct_formula}), we need only examine the computational time, which can be decomposed as $T(\gammab) = t(\gammab)\left(\nb\right)$, where $t(\gammab)$ is the average computation time per batch, multiplied by the number of batches.
Warm-starting an algorithm may be used to decrease the computation time per batch $t(\gammab)$ for $\gammab < 1$, leading to a possible increase in computational efficiency.

To investigate computational efficiency, we assumed that computation time per batch can be approximated as $t(\gammab) \propto \gammab^p$ for some $p \geq 0$.
Results for some values of $p$ are shown in Figure \ref{fig:efficiency_theory}.
Efficiency gains increase with larger exponent $p$, but are positive even for small values of $p$ when $\gammab$ is sufficiently close to 1.
At $p=1$, i.e., for problems where computational time is proportional to the number of samples replaced in the batch, computational efficiency is maximized with maximally overlapping batches of \cite{Meketon1984}.
When $p$ is $0.71$, on the other hand, computational efficiency is observed for values of $\gammab$ from slightly lower than $0.3$ to $1$, and efficiency is maximized at a relatively flat region between $\gammab=0.5$ and $\gammab=0.8$.

We tested this theoretical relationship using the Newsvendor problem solved with insertion sort.
Insertion sort is an efficient sorting algorithm when the values are already largely sorted. 
Therefore, as overlapping increases, insertion sort results in a decreased solution time per batch for the NVD problem. 
The average computation time per batch was found empirically to be well approximated by power law with $p = 0.71$.
The empirically measured computational efficiency is compared to the theoretical result in Figure \ref{fig:efficiency_nv}, where the theoretical computational efficiency, shown in solid line, is repeated from Figure \ref{fig:efficiency_theory}.
It appears that empirical observations match with the theoretical efficiency calculations well.

In cases where the exponent $p$ can be estimated, the above theoretical method may be used to determine a reasonable value for $\gammab$.
When warm-starting can be used to solve the batch sampling problems, we recommend the users to perform preliminary computations to determine how computational time per batch changes as $\gammab$ changes. 
We can see in Figure \ref{fig:efficiency} that for some values of $p$, efficiency gain is nearly flat over a region of $\gammab$.
In this case, we recommend choosing a $\gammab$ near the right side of this region.
This choice keeps efficiency high while decreasing computation time and preventing large decreases in efficiency in the event that $p$ was overestimated.
For example, in the case of NVD, we recommend the use of value $\gammab=0.8$ as this value provides a high computational efficiency with minimal overall computation time across values of $\gammab$ with high efficiency.

@@

A final question on computational efficiency would be whether OMRP is more efficient than nonoverlapping MRP with a larger number of batches.
When generating observations is cheap, one could potentially increase the number of batches in MRP to obtain estimators with lower variances. 
In the nonoverlapping case, $\var{VG} \propto \frac{1}{k}$, so decreasing $\var{VG}$ to the level of maximally overlapping batches requires a $50\%$ increase in the number of batches, and thus a $50\%$ increase in computation time.
Continuing on in this way for various values of $\gammab$, we obtain the solid black curve in Figure \ref{fig:comp_work_comparison}. 
This curve shows the increase in computational effort due to an increased number of batches in MRP in order to obtain the same variance level for OMRP across $\gammab$.
If the work increase by OMRP is less than this solid black curve, then OMRP is computationally more efficient than increasing the number of batches in MRP. 
Again, assuming average computation time per batch can be represented as $T(\gammab) \propto \gammab^p$, Figure \ref{fig:comp_work_comparison} compares the increase in computational work for OMRP for several values of $p$ with the work increase for MRP to get an equivalent decrease in variance.
We again see that OMRP is advantageous for larger values of $p$.
For small values of $p$, some efficiency gain is observed at $\gammab$ values close to 1.
